\documentclass{article}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{biblatex}
\usepackage{hyperref}
\addbibresource{main.bib}


\title{The Vyom Hypothesis}
\author{Siddarth S. Nair}
\date{May 2025}

\begin{document}
\begin{titlepage}
    \centering
    \vspace*{\fill}
    {\Huge\bfseries The Vyom Hypothesis \par}
    \vspace{1cm}
    {\Large Siddarth S. Nair \par}
    \vspace{0.5cm}
    {\large May 2025 \par}
    \vspace*{\fill}
    \thispagestyle{empty}
\end{titlepage}



\newpage
\begin{abstract}
    The universe came from nothing, and thus, it must return to nothing. While modern physics, including quantum mechanics and general relativity, has made significant progress in explaining how systems behave, this paper offers a speculative framework to explore why systems behave the way they do. It introduces the postulate that all systems follow a singular principle: the tendency to become non-existent. By defining existence as a function of a system's physical, electromagnetic, and quantum complexity, this hypothesis aims to provide insight into the nature of fundamental forces, the role of dark energy, and the possibility of a fractal universe structure. Ultimately, the theory aspires to be a step towards the theory of everything.
\end{abstract}



\newpage
\tableofcontents



\newpage
\section{Introduction}


\subsection{Backstory}
This paper originated from a narrative challenge that the author encountered while working on a now-shelved science fiction story. A critical plot event lacked a plausible explanation within the bounds of current physics theories. Rather than relying on narrative conveniences or abandoning the concept outright, he sought to develop a theoretical explanation of his own. This initial spark eventually spiraled into the making of this paper.


\subsection{Motivations}
Modern physics has made a significant development in explaining how systems behave via the standard model, quantum mechanics \cite{dirac1930} and general relativity \cite{einstein1915}. However, these theories exist in isolated patches, and also fail to explain why systems behave the way they do. This is where The Vyom Hypothesis comes in, building upon established theories, it provides a speculative answer on why the fundamental forces exist, why the expansion of space is accelerating and the fate of the universe. 


\subsection{Shortcomings}
As the author is currently a high school student, this paper reflects a developing understanding of advanced physics. While the Vyom Hypothesis proposes a unifying framework across the physical, electromagnetic, and quantum domains, many aspects remain unexplained, speculative and incomplete. Further education and years of study will be necessary to rigorously formulate, test, and potentially validate this theory. This paper and all that it entails is intended as a foundational exploration rather than a finalized model.



\newpage
\section{Postulate}

\begin{center}
Every system tends to non-existence.
\end{center}

\subsection{Existence}
Existence is the measure of a system’s ability to interact across three fundamental domains: physical, electromagnetic, and quantum. The existence of a system is determined by the combined complexity it holds in each of these three domains. Therefore, to achieve true non-existence, a system must reduce its complexity in all three domains to zero.

This tendency can be intrinsic, where the system moves toward non-existence through its own internal behavior, or extrinsic, where it becomes non-existence due to external influences. Importantly, this postulate does not imply that existence must strictly decrease. Temporary increases in domain specific or overall complexity can occur, for instance, during the formation of stars. But over long timescales, the net trend of systems remains toward non-existence.

As far as current observations suggest, all systems follow one of two primary pathways toward nonexistence: attraction or repulsion. A third possibility, remaining in complete stasis, corresponds to a state of true nonexistence, where no interaction or change occurs. All fundamental forces, then, can be interpreted as manifestations of a system's intrinsic tendency to reduce its existence.

\subsection{Terminology}
\begin{itemize}
    \item \textbf{Non-existence}: A state in which a system cannot interact with external systems or its internal subsystems.
    \item \textbf{Tendency}: An abbreviation for the phrase “tendency to become non-existent.” It also refers to the pathway a system takes in its pursuit of non-existence, such as through attraction or repulsion.
    \item \textbf{Attraction}: A pathway of tendency in which subsystems move closer together.
    \item \textbf{Repulsion}: A pathway of tendency in which subsystems move further apart.
    \item \textbf{Domain}: Refers to the three fundamental fields of interaction: physical, electromagnetic, or quantum.
    \item \textbf{Complexity}: The magnitude of possible interactions within a specific domain.
    \item \textbf{PH complexity}: Abbreviation of “physical complexity.”
    \item \textbf{EM complexity}: Abbreviation of “electromagnetic complexity.”
    \item \textbf{QU complexity}: Abbreviation of “quantum complexity.”
\end{itemize}

\subsection{Physical Complexity}
Physical complexity is defined as the number of realistic configurations a system can occupy when isolated from external influences. Imagine placing the system inside an opaque box and observing it at random intervals, its internal arrangement might differ each time due to inertial motion, thermodynamic activity, or internal forces. This variability reflects the system’s inherent structural complexity.

To reduce PH complexity, a system must condense into an extremely dense state, thereby eliminating the degrees of freedom of its constituent particles. Such a configuration has minimal, ideally zero, structural variability. Gravity is the curvature of spacetime as described by general relativity \cite{einstein1915}, but it can also be interpreted as a manifestation of the intrinsic tendency to reduce PH complexity. 

For regular matter, spacetime curves around mass not arbitrarily, but because matter follows its path toward nonexistence, and gravity is the mechanism through which that reduction in physical degrees of freedom is realized.

However, extrinsic pathways to near-zero PH complexity also exist. For example, during the heat death of the universe, particles become so widely separated by cosmic expansion that they effectively behave as isolated systems. Though still in motion, these particles interact minimally and exhibit no internal degrees of freedom, resulting in negligible PH complexity. The strong nuclear force also acts as an extrinsic tendency, as will be discussed later

\subsection{Electromagnetic Complexity}
Electromagnetic complexity is defined as the number of unique electromagnetic interactions that exist within a system. It arises from the presence and arrangement of electric charges. Just as systems tend to reduce PH complexity by collapsing into simpler structures, they reduce EM complexity by minimizing net charge imbalances.

Electromagnetic attraction and repulsion are manifestations of a system's tendency to minimize EM complexity. Opposite charges, one with an excess of electrons and the other with a deficit, are drawn together in order to redistribute charges and achieve mutual neutrality. This attraction represents the system's attempt to reduce net charge imbalance, thereby lowering EM complexity. Conversely, like charges both possess either an excess or a deficiency of electrons. Bringing them closer would increase the system's imbalance. Thus, they repel, moving apart until their mutual interaction becomes negligible, which again results in lower EM complexity.

\subsection{Quantum Complexity}
Quantum complexity is the inherent randomness associated with a system due to the probabilistic nature particles at the quantum level, most notably described by Heisenberg’s uncertainty principle \cite{heisenberg1927}. This randomness arises from the fact that the properties of particles, like position and momentum, cannot be simultaneously known with perfect precision, which introduces fundamental uncertainty into their behavior.

Achieving QU non-existence involves reducing the inherent randomness in a system. Particles either tend to move apart to minimize internal influence, effectively becoming isolated subsystems, or they clump together tightly, reducing net uncertainty through confinement. These tendencies manifest as the weak and strong nuclear forces, respectively. It can thus be inferred that the strong nuclear force is a manifestation of the system’s drive to reduce both PH and QU complexity. However, even in complete isolation, each particle still carries an intrinsic uncertainty. Therefore, to truly reach quantum non-existence, these particles must cease to exist altogether. Reducing quantum complexity, then, becomes the final hurdle in achieving true non-existence.



\newpage
\section{Case Studies}


\subsection{The Universe}

\subsubsection{Matter}
Matter, as a system, possesses significant complexity in all three domains. While all three complexities exist at every scale, the dominant complexity, and thus the system’s net tendency toward non-existence, depends on the scale of observation.
\begin{itemize}
    \item At cosmic scales, gravitational attraction is the dominant force. It reduces PH complexity by causing matter to clump together, forming stars, galaxies, and ultimately, extremely dense objects such as black holes. Once matter reaches this state, the system loses nearly all degrees of freedom, eliminating PH complexity, and its electromagnetic interactions become negligible. At this point, the object is governed almost entirely by quantum complexity. The only remaining pathway toward non-existence is through quantum decay processes, namely, Hawking radiation \cite{hawking1975}.
    \item At macroscopic scales, such as in objects we interact with daily, both PH and EM complexity govern structure and interactions, such as planetary gravity and electromagnetic forces.
    \item At molecular and atomic scales, EM and QU complexities dominate as hadrons and electrons bind themselves into atoms via the electromagnetic and strong force respectively.
    \item At subatomic and quantum scales, QU complexity dominates. The probabilistic nature of particles, governed by the uncertainty principle, dictates behavior and interactions, such as the strong and weak nuclear forces.
\end{itemize}

\subsubsection{Spacetime}
Spacetime, by contrast, exists solely in the QU domain. It is saturated with quantum fluctuations arising from its uncertainty, spontaneous appearances and annihilation of particle-antiparticle pairs, which contribute to its QU complexity. In order to reduce it, spacetime expands. However, uniform expansion is not enough to fully suppress these fluctuations as they annihilate faster than the expansion rate. As a result, the expansion of spacetime accelerates. When this acceleration eventually causes the relative speed of expansion between nearby points to exceed the speed of light, quantum fluctuations can no longer annihilate. At that threshold point, the final fluctuation becomes a pair of real, isolated particles, marking the end of QU complexity in spacetime.

\subsubsection{The Cosmos}
The universe as a whole is governed by the combined tendencies of matter and spacetime to move toward non-existence. However, since spacetime possesses a higher magnitude of existence, it dominates the universe's net tendency. As a result, even matter is passively driven apart by the expanding fabric of spacetime. This acts as an extrinsic pathway to non-existence, where matter becomes so widely dispersed that meaningful interactions between particles cease. Over time, as the expansion accelerates, it will overcome all binding forces, eventually tearing apart even hadrons, leaving behind a sea of fundamental particles, quarks and leptons, in isolation. This represents the universe's final, inert state.

Speculatively, this may not mark the absolute end. The remaining particles might decay further, transferring their energy into the quantum vacuum as dark energy. This accumulated energy could then fuel the emergence of smaller, fractal universes, an idea explored in The Fractal Universe Hypothesis, presented later in the paper.


\subsection{The Standard Model}

The Standard Model describes the fundamental building blocks of matter and the interactions that govern them \cite{standardmodel}. Each component of the model follows its own tendency toward non-existence through distinct decay or annihilation pathways.

\subsubsection{Neutrons}
Neutrons present in atomic nuclei are in a pseudo-stable, near-non-existent state. They condense into nuclei via the strong nuclear force to reduce both PH complexity and QM complexity. However, when the extrinsic accelerated expansion of the universe overcomes nuclear binding forces, nuclei dissociate into isolated free hadrons. A free neutron has a half-life of approximately $611 \pm 1$ seconds. Through beta decay, it transforms into a proton, electron, electron antineutrino, and very rarely, emits a gamma ray.

\subsubsection{Protons}
Currently, protons are considered stable within the Standard Model. However, some Grand Unified Theories predict that protons may decay after around $10^{34}$ years. In such scenarios, a proton would decay into a positron and a neutral pion. The latter quickly decays into two gamma rays, which then follow their own tendency.

\subsubsection{Annihilation}
Every particle in the Standard Model has a corresponding antiparticle, such as the electron and the positron. When these particles encounter each other, they annihilate, converting their mass into energy and consequent byproducts. This annihilation process completely reduces EM complexity, QM complexity, and net complexity, leading to true non-existence. For example, a positron formed via proton decay may annihilate with an electron, producing two photons.

\subsubsection{Electromagnetic Waves}
Photons and gamma rays, forms of electromagnetic radiation, follow their tendency through the extrinsic influence of spacetime’s expansion. As they propagate through expanding space, their wavelengths increase, causing them to lose energy. This phenomenon, known as cosmological redshift, results in a gradual decrease in their EM and QM complexity, eventually rendering them completely non-existent \cite{hubble1929}.

\subsubsection{Extrinsic Tendency via Spacetime Expansion}
The accelerated expansion of spacetime exerts an extrinsic tendency on all particles. Over time, it isolates particles from one another by stretching the space between them faster than they can interact. Once particles become sufficiently separated, they can no longer engage in decay, annihilation, or bonding processes. This extrinsic isolation halts any internal mechanisms of complexity reduction, leaving the particle effectively "frozen" in its current state. In the final stages of the universe, this dispersion will reduce both EM and PH complexity to near zero, with only the residual QM complexity of isolated fundamental particles remaining.

\subsubsection{Atoms and Molecules}
Free hadrons, such as protons and neutrons, attract each other via gravity to reduce their PH complexity and eventually fuse under extreme conditions, such as in stars, with the help of the strong nuclear force. These fused nuclei, forming atoms, naturally attract electrons to achieve near net neutrality, thereby reducing their EM complexity.

However, in many atoms—particularly those that are not noble gases, there remains abnormally high QU complexity due to unfulfilled electron orbitals. These unfilled orbitals allow for multiple potential configurations of electron positions, increasing quantum uncertainty. To reduce this QU complexity, such atoms tend to gain, lose, or share electrons through chemical reactions. This process can sometimes result in an increase in EM complexity, but the overall drive is toward lowering total domain complexity.



\subsection{Entropy}
Entropy is the inherent randomness of a system, and increases with every process that occurs. According to thermodynamics, the universe moves towards a state of maximal entropy, the aptly named, Heat Death. Initially, this seems to contradict with one of the central claim's of this paper, that every system tends to reduce its PH complexity.

However, there is an important distinction between the two. While both of them describe a system's configurational behavior, they do so in opposite terms. Entropy measures the number of micro-states a system can occupy in a macro-state, it is a measure of disorder or unpredictability. Physical complexity, as defined, refers to the number of realistic configurations a system can meaningfully maintain. 

Take for example a star dying and turning into a black hole. This process increases entropy as there is more disorder, while at the same time, it decreases PH complexity, as the particles lose all degrees of freedom. Thus, entropy and PH complexity are two sides of the same coin.



\newpage
\section{Speculative Ideas}


\subsection{Fractal Universes}

Since the expansion of spacetime is accelerating with time, there will come a point in the far future when the speed of expansion exceeds the speed of light at quantum scales. When this moment arrives, hereby referred to as the Vyom Point, virtual particles arising from quantum fluctuations will fail to annihilate. They will become a pair of real particles, drifting away from each other in isolated subsystems. These particles will have zero PH and EM complexity but will still possess QU complexity due to Heisenberg’s uncertainty principle.

At the Vyom Point, the virtual particles become real and isolated, as described above. With no other pathway available to fulfill their tendency toward non-existence, they resort to initiating a new Big Bang event. Drawing energy from spacetime, they trigger the birth of a new universe. Since an innumerable number of quantum fluctuations disintegrate at the same time, these new universes emerge as fractal universes, each isolated from the others.

However, this process is not 100\% efficient, as dictated by the second law of thermodynamics \cite{newton1687, clausius1865}. Each Big Bang increases entropy locally, that is, within the new universe. But to enable this, entropy must be pushed outside the universe. The problem is that outside the universe lies nothing. True non-existence cannot become more disordered, as it has no degrees of freedom. It does not exist. Therefore, the entropy, along with its associated energy, is lost.

Over many generations of fractal universes, the total energy originating from the first universe is gradually depleted. Eventually, all energy is gone. True non-existence is finally reached.



\newpage
\printbibliography

\end{document}